The Design and Implementation of the Synchronous Language \CEU

### 2016-03-28 paper.tex
...

We decided for the unique and queue-based semantics for the reasons that 
follow:
%
\begin{itemize}
\item \textbf{A ``tick'' is too abstract and imprecise:}
Outside the domain of hardware specification, a tick has no natural counterpart 
in the real world.
Also, since ticks require no time regularity~\cite{esterel.multiclock}, the two 
approaches for Esterel in Figure~\ref{fig.ticks} are legitimate, but lead to 
different behaviors for the same sequence of inputs.
%
\item \textbf{Events are never simultaneous:}
From a rigorous point of view, event occurrences are infinitesimal, having zero 
probability of being simultaneous.
This way, we believe that the notion of simultaneity should not be imposed by 
the language, but explicitly defined for each use case.
In the end of the section, we present a specification to detect 
``simultaneous'' button clicks.
%
Note that in the case of Esterel, simultaneity depends on the imprecise 
definition of ticks.
For instance, in Figure~\ref{fig.ticks}, the events \code{B} and \code{C} are
simultaneous, even though \code{A} and \code{B} actually happen much closer to 
one another.
\item \textbf{Unique input events imply mutual exclusion:}
Because they are atomic, reactions to different events never overlap.
The possibility to reason about each input individually is a prerequisite for 
the temporal analysis to be discussed in Section~\ref{sec.ceu.analysis}.
\end{itemize}

\begin{figure}
\begin{minipage}[t]{0.57\linewidth}
\begin{lstlisting}[numbers=left,xleftmargin=3em,mathescape=true]
input void LEFT_CLICK;
input void RIGHT_CLICK;
event void middle_click;
loop do
    par/or do
        AWAIT_AND(LEFT_CLICK, RIGHT_CLICK);
        emit middle_click;
    with
        AWAIT_OR(LEFT_CLICK, RIGHT_CLICK);
        await 200 ms;
    end
end
\end{lstlisting}
\end{minipage}
%
\begin{minipage}[t]{0.43\linewidth}
\begin{lstlisting}[numbers=left,xleftmargin=3em,firstnumber=13]
#define AWAIT_AND(e1, e2)   \
    par/and do              \
        await e1;           \
    with                    \
        await e2;           \
    end
#define AWAIT_OR(e1, e2)    \
    par/or do               \
        await e1;           \
    with                    \
        await e2;           \
    end
\end{lstlisting}
\end{minipage}
%\rule{8.4cm}{0.37pt}
\caption{ Application defines that a \code{middle\_click} event occurs whenever 
both \code{LEFT\_CLICK} and \code{RIGHT\_CLICK} occur within 200 milliseconds.
\label{lst.simultaneous}
}
\end{figure}

Figure~\ref{lst.simultaneous} emulates a \code{milddle\_click} event (ln. 3) in 
terms of \code{LEFT\_CLICK} and \code{RIGHT\_CLICK} (ln. 1--2).
If both events occur, we emit the internal event \code{middle\_click} (ln.  
6--7).
However, if one of them occurs and the \emph{200ms} timer expires (ln. 9--10), 
we abort the whole behavior with the \code{par/or} (ln. 5--11) and try again 
with the enclosing \code{loop} (ln. 4).
%
We discuss internal events in Section~\ref{sec.ceu.ints} and timers in 
Section~\ref{sec.ceu.timers}.
%
The macros \code{AWAIT\_AND} (ln. 13--18) and \code{AWAIT\_OR} (ln. 19--24) are 
simple expansions to a \code{par/and} and \code{par/or} for better readability.
%
In this specification, ``simultaneous'' means ``within 200 milliseconds'', 
which is a huge amount of time for a language-defined tick, and which would 
break the synchronous hypothesis.
%
Note that a similar implementation for Esterel would not rely on the tick 
notion of simultaneity either.
%
Note also that the application in \CEU could define an external input 
\code{TICK} and change the expiring event from \code{200ms} to \code{TICK} (ln.  
10) to recover the abstract notion of simultaneity.

The synchronous hypothesis for \CEU holds if the reactions run faster than the 
rate of incoming input events.
Otherwise, the system will accumulate delays between the real occurrence and 
actual reaction to events.
%
This is also the case for the variable-length-tick approach of Esterel, since 
the more inputs to handle, the longer the reaction takes, and the more inputs 
will accumulate for subsequent ticks.
%
For the fixed-length-tick approach of Esterel, a breach in the synchronous 
hypothesis causes timing violations, which can be avoided with \emph{worst case 
reaction time} analysis to infer an appropriate value for the tick 
length~\cite{esterel.fixed}.

@@

Esterel makes no semantic distinctions between internal and external signals.
In particular, programs can emit different external input signals 
simultaneously, with all coexisting during a reaction.
%
In \CEU, a reaction starts from an external input event and programs cannot 
emit inputs at all.
Therefore, the occurring input event is unique during the entire reaction, 
resulting in intrinsic queue-based handling.
%
In contrast, programs can emit internal events but these follow a stack-based 
execution policy, similar to subroutine calls in typical programming languages.
%
Figure~\ref{lst.prints} illustrates the use of internal signals (events) in 
Esterel (a) and \CEU (b).
%
In Esterel, when \code{A} occurs, \code{B} is emitted (ln. 4--5) and both 
events become active, resulting in the invocation of \code{f()} and \code{g()} 
in no particular order.
%
In \CEU, the occurrence of \code{A} makes the program behave as follows:
%
{\small
\begin{enumerate}
\setlength{\itemsep}{0pt}
\item 1st trail awakes (ln. 4), emits \code{b}, and pauses.
\item 2nd trail awakes (ln. 8), calls \code{\_g()}, and terminates.
\item 1st trail (on top of the stack) resumes, calls \code{\_f()}, and 
    terminates.
\item Both trails have terminated, so the \code{par/and} rejoins, and the 
program also terminates.
\end{enumerate}
}

\begin{figure}
\begin{minipage}[t]{0.43\linewidth}
\begin{lstlisting}[numbers=left,xleftmargin=3.5em,mathescape=true]
input A;    // external
signal B;   // internal
[[
    await A;
    emit B;
    call f();
$\|$
    await B;
    call g();
]]
\end{lstlisting}
\centering\small{(a) Esterel}
\end{minipage}
%
\begin{minipage}[t]{0.53\linewidth}
\begin{lstlisting}[numbers=left,xleftmargin=3.1em]
input void A;  // external (in uppercase)
event void b;  // internal (in lowercase)
par/and do
    await A;
    emit b;
    _f();
with
    await b;
    _g();
end
\end{lstlisting}
\centering\small{(b) \CEU}
\end{minipage}
%\rule{8.4cm}{0.37pt}
\caption{ Internal signals (events) in Esterel and \CEU: similar syntax, but 
different semantics.
\label{lst.prints}
}
\end{figure}

@@

@@

\begin{figure}
\begin{lstlisting}[numbers=left,xleftmargin=2em]
event void e,f;
loop do
    par/or do
        await e;
    with
        emit e;     // w/o delayed awaits, the emit awakes 1st trail
        await f;    // and restarts the loop instantaneously
    end
end
\end{lstlisting}
\caption{ Delayed awaits prevents re-execution of statements by design.
\label{lst.await}
}
\end{figure}

@@

\begin{figure}[t]
%\begin{minipage}[t]{0.50\linewidth}
\begin{lstlisting}[numbers=left,xleftmargin=3em]
event void e;
par do
    loop do
        <...>     // code that awaits some period
        emit e;   // periodic request
    end
with
    loop do
        <...>     // code to execute immediately and then periodically
        await e;  // await after
    end
end
\end{lstlisting}
%\end{minipage}
%
%\begin{minipage}[t]{0.50\linewidth}
%\begin{lstlisting}[numbers=left,xleftmargin=3em]
%event void e;
%var bool should_execute = false;
%par do
%    loop do
%        <...>         // code that awaits
%        if <...> then // some conditon
%            should_execute = true;
%            emit e;
%        end
%    end
%with
%    loop do
%        if should_execute then
%            <...>     // code to execute
%        end
%        await e;
%    end
%end
%\end{lstlisting}
%\end{minipage}
%
\caption{ An example that circumvents the \emph{delayed await} by post-fixing 
the \code{await} inside the \code{loop}.
% (in the left), and by copying the condition test (in the right).
\label{lst.delay}
}
\end{figure}

@@

\begin{figure}[b]
\begin{minipage}[t]{0.50\linewidth}
\begin{lstlisting}[numbers=left,xleftmargin=3.5em]
native do
    #define NUM 10
    void f  (void)  { <...> }
    void g  (int v) { <...> }
    int  id (int v) { <...> }
end
par/and do
    _f();
with
    _g(_id(_NUM));
end
\end{lstlisting}
\centering\small{(a)}
\end{minipage}
%
\begin{minipage}[t]{0.44\linewidth}
\begin{lstlisting}
native @const _NUM;
native @pure  _id;
native @safe  _f with _g;

@@

@@

\begin{figure}[t]
\begin{minipage}[t]{0.31\linewidth}
\begin{lstlisting}
input void A;
event void e;
// TRAIL 0 - lbl Main
par/and do
  // TRAIL 0 - lbl Main
  await e;
  // TRAIL 0 - lbl Awake_e
  // TRAIL 0 - lbl And_chk
with
  // TRAIL 1 - lbl And_sub_2
  await A;
  // TRAIL 1 - lbl Awake_A_1
  emit e;
  // TRAIL 1 - lbl Emit_cont
  // TRAIL 1 - lbl And_chk
end
// TRAIL 0 - lbl And_out
await A;
// TRAIL 0 - lbl Awake_A_2

@@

The event-driven approach of \CEU is widespread~\cite{sync_async.whynotthreads} 
and popular in many software communities, such as web frameworks (e.g., 
\emph{jQuery}~\cite{js.jquery} and \emph{Node.js}~\cite{js.node}), GUI toolkits 
(e.g., \emph{Tcl/Tk}~\cite{tcl.tk} and \emph{Java Swing}~\cite{java.swing}), 
and Games~\cite{gamepatterns}.
%
Like \CEU, event-driven programming is essentially synchronous, i.e., events go 
through a queue and are dispatched sequentially and atomically to prevent race 
conditions.
%
We believe that for software design, this approach is more familiar to 
programmers and simplifies the reasoning about concurrency.
For instance, the uniqueness of external events in \CEU is a prerequisite for 
the temporal analysis that enables safe shared-memory concurrency.

>>>>

We decided for the unique and queue-based semantics for the reasons that 
follow:
%
\begin{itemize}
\item \textbf{A ``tick'' is too abstract and imprecise:}
Outside the domain of hardware specification, a tick has no natural counterpart 
in the real world.
Also, since ticks require no time regularity~\cite{esterel.multiclock}, the two 
approaches for Esterel in Figure~\ref{fig.ticks} are legitimate, but lead to 
different behaviors for the same sequence of inputs.
%
\item \textbf{Events are never simultaneous:}
From a rigorous point of view, event occurrences are infinitesimal, having zero 
probability of being simultaneous.
This way, we believe that the notion of simultaneity should not be imposed by 
the language, but explicitly defined for each use case.
Note that in the case of Esterel, simultaneity depends on the imprecise 
definition of ticks.
For instance, in Figure~\ref{fig.ticks}, the events \code{B} and \code{C} are
simultaneous, even though \code{A} and \code{B} actually happen much closer to 
one another.
After introducing internal events and first-class timers, we show an example 
that detects simultaneous button clicks in Section~\ref{sec.ceu.discussion}.
%
\item \textbf{Unique input events imply mutual exclusion:}
Because they are atomic, reactions to different events never overlap.
The possibility to reason about each input individually is a prerequisite for 
the temporal analysis to be discussed in Section~\ref{sec.ceu.analysis}.
\end{itemize}

The synchronous hypothesis for \CEU holds if the reactions run faster than the 
rate of incoming input events.
Otherwise, the system will accumulate delays between the real occurrence and 
actual reaction to events.
%
This is also the case for the variable-length-tick approach of Esterel, since 
the more inputs to handle, the longer the reaction takes, and the more inputs 
will accumulate for subsequent ticks.
%
For the fixed-length-tick approach of Esterel, a breach in the synchronous 
hypothesis causes timing violations, which can be avoided with \emph{worst case 
reaction time} analysis to infer an appropriate value for the tick 
length~\cite{esterel.fixed}.
% TODO: Such analsysis is orthogonal

@@

Esterel makes no semantic distinctions between internal and external signals.
In particular, programs can emit external and internal signals simultaneously, 
with all coexisting during the entire reaction.
%
%In \CEU, a reaction starts from an external input event and programs cannot 
%emit inputs at all.
%Therefore, the occurring input event is unique during the entire reaction, 
%resulting in intrinsic queue-based handling.
%
In \CEU, subsequent external events coming from the environment define a 
timeline, which programs cannot interfere by emitting new external events.
%
Howerver, \CEU supports internal events which programs can \code{emit} in 
addition to \code{await}.
In contrast with external events, internal events follow a stack-based 
execution policy similar to subroutine calls in typical programming languages.
%
Figure~\ref{lst.prints} illustrates the use of internal signals (events) in 
Esterel (a) and \CEU (b).
%
In Esterel, when \code{A} occurs, \code{B} is emitted (ln. 4--5) and both 
events become active, resulting in the invocation of \code{f()} and \code{g()} 
in no particular order.
%
In \CEU, the occurrence of \code{A} makes the program behave as follows:
%
{\small
\begin{enumerate}
\setlength{\itemsep}{0pt}
\item 1st trail awakes (ln. 4), emits \code{b}, and pauses.
\item 2nd trail awakes (ln. 8), calls \code{\_g()}, and terminates.
\item 1st trail (on top of the stack) resumes, calls \code{\_f()}, and 
    terminates.
\item Both trails have terminated, so the \code{par/and} rejoins, and the 
program also terminates.
\end{enumerate}
}

\begin{figure}[b]
\begin{minipage}[t]{0.43\linewidth}
\begin{lstlisting}[numbers=left,xleftmargin=3.5em,mathescape=true]
input A;    // external
signal B;   // internal
[[
    await A;
    emit B;
    call f();
$\|$
    await B;
    call g();
]]
\end{lstlisting}
\centering\small{(a) Esterel}
\end{minipage}
%
\begin{minipage}[t]{0.53\linewidth}
\begin{lstlisting}[numbers=left,xleftmargin=3.1em]
input void A;  // external (in uppercase)
event void b;  // internal (in lowercase)
par/and do
    await A;
    emit b;
    _f();
with
    await b;
    _g();
end
\end{lstlisting}
\centering\small{(b) \CEU}
\end{minipage}
%\rule{8.4cm}{0.37pt}
\caption{ Internal signals (events) in Esterel and \CEU: similar syntax, but 
different semantics.
\label{lst.prints}
}
\end{figure}

@@

\begin{figure}[b]
\begin{lstlisting}[numbers=left,xleftmargin=2em]
event void e,f;
loop do
    par/or do
        await e;
    with
        emit e;     // w/o delayed awaits, the emit awakes 1st trail
        await f;    // and restarts the loop instantaneously
    end
end
\end{lstlisting}
\caption{ Delayed awaits prevents re-execution of statements by design.
\label{lst.await}
}
\end{figure}

@@

\begin{figure}
%\begin{minipage}[t]{0.50\linewidth}
\begin{lstlisting}[numbers=left,xleftmargin=3em]
event void e;
par do
    loop do
        <...>     // code that awaits some period
        emit e;   // periodic request
    end
with
    loop do
        <...>     // code to execute immediately and then periodically
        await e;  // await after
    end
end
\end{lstlisting}
%\end{minipage}
%
%\begin{minipage}[t]{0.50\linewidth}
%\begin{lstlisting}[numbers=left,xleftmargin=3em]
%event void e;
%var bool should_execute = false;
%par do
%    loop do
%        <...>         // code that awaits
%        if <...> then // some conditon
%            should_execute = true;
%            emit e;
%        end
%    end
%with
%    loop do
%        if should_execute then
%            <...>     // code to execute
%        end
%        await e;
%    end
%end
%\end{lstlisting}
%\end{minipage}
%
\caption{ An example that circumvents the \emph{delayed await} by post-fixing 
the \code{await} inside the \code{loop}.
% (in the left), and by copying the condition test (in the right).
\label{lst.delay}
}
\end{figure}

@@

@@

\begin{figure}
\begin{minipage}[t]{0.50\linewidth}
\begin{lstlisting}[numbers=left,xleftmargin=3.5em]
native do
    #define NUM 10
    void f  (void)  { <...> }
    void g  (int v) { <...> }
    int  id (int v) { <...> }
end
par/and do
    _f();
with
    _g(_id(_NUM));
end
\end{lstlisting}
\centering\small{(a)}
\end{minipage}
%
\begin{minipage}[t]{0.44\linewidth}
\begin{lstlisting}
native @const _NUM;
native @pure  _id;
native @safe  _f with _g;

@@

\CEU aims to offer a simpler semantics than Esterel while keeping the power of 
synchronous languages, with a tractable concurrency model, reactive and 
deterministic execution, and safe support for orthogonal abortion of lines of 
execution.

The notion of time based on events instead of ticks 
(Section~\ref{sec.ceu.exts}) is widespread~\cite{sync_async.whynotthreads} and 
popular in many software communities, such as web frameworks (e.g., 
\emph{jQuery}~\cite{js.jquery} and \emph{Node.js}~\cite{js.node}), GUI toolkits 
(e.g., \emph{Tcl/Tk}~\cite{tcl.tk} and \emph{Java Swing}~\cite{java.swing}), 
and Games~\cite{gamepatterns}.
%
Like \CEU, event-driven programming is essentially synchronous, i.e., events go 
through a queue and are dispatched sequentially and atomically to prevent race 
conditions.
%
We believe that for software design, this approach is more familiar to 
programmers and simplifies the reasoning about concurrency.
For instance, the uniqueness of external events in \CEU is a prerequisite for 
the temporal analysis for concurrent statements.

With the event-based semantics, \CEU does not impose a notion of simultaneity 
for events, requiring each application to define it only when necessary.
%
Figure~\ref{lst.simultaneous} emulates a \code{milddle\_click} event (ln. 3) in 
terms of ``simultaneous'' occurrences of \code{LEFT\_CLICK} and 
\code{RIGHT\_CLICK} (ln. 1--2).
If both events occur, we emit the internal event \code{middle\_click} (ln.  
6--7).
However, if one of them occurs and the \emph{200ms} timer expires (ln. 9--10), 
we abort the whole behavior with the \code{par/or} (ln. 5--11) and try again 
with the enclosing \code{loop} (ln. 4).
%
In this specification, ``simultaneous'' means ``within 200 milliseconds'', 
which is a huge amount of time for a language-defined tick, and which would 
break the synchronous hypothesis.
%
Note that a similar implementation for Esterel would not rely on the tick 
notion of simultaneity either.
%
Note also that the application in \CEU could define an external input 
\code{TICK} and change the expiring event from \code{200ms} to \code{TICK} (ln.  
10) to recover the abstract notion of simultaneity.

\begin{figure}
\begin{minipage}[t]{0.57\linewidth}
\begin{lstlisting}[numbers=left,xleftmargin=3em,mathescape=true]
input void LEFT_CLICK;
input void RIGHT_CLICK;
event void middle_click;
loop do
    par/or do
        AWAIT_AND(LEFT_CLICK, RIGHT_CLICK);
        emit middle_click;
    with
        AWAIT_OR(LEFT_CLICK, RIGHT_CLICK);
        await 200 ms;
    end
end
\end{lstlisting}
\end{minipage}
%
\begin{minipage}[t]{0.43\linewidth}
\begin{lstlisting}[numbers=left,xleftmargin=3em,firstnumber=13]
#define AWAIT_AND(e1, e2)   \
    par/and do              \
        await e1;           \
    with                    \
        await e2;           \
    end
#define AWAIT_OR(e1, e2)    \
    par/or do               \
        await e1;           \
    with                    \
        await e2;           \
    end
\end{lstlisting}
\end{minipage}
%\rule{8.4cm}{0.37pt}
\caption{ Application defines that a \code{middle\_click} event occurs whenever 
both \code{LEFT\_CLICK} and \code{RIGHT\_CLICK} occur within 200 milliseconds.
%
The macros \code{AWAIT\_AND} (ln. 13--18) and \code{AWAIT\_OR} (ln. 19--24) are 
simple expansions to a \code{par/and} and \code{par/or} for better readability.
\label{lst.simultaneous}
}
\end{figure}

internal events
    - flexible, coroutines, resumable exceptions
    - yet simple, programmers understand stacks
    - most of the time doesn't matter

Intra-reaction determinism is another
The temporal analsysi (Section~\ref{sec.ceu.analysis}).

The temporal analsysi (Section~\ref{sec.ceu.analysis}).
simple, programmers understand

targets control-intensive applications and provides abstractions that can 
express program flow specifications concisely.

@@

\begin{figure}[b]
\begin{minipage}[t]{0.31\linewidth}
\begin{lstlisting}
input void A;
event void e;
// TRAIL 0 - lbl Main
par/and do
  // TRAIL 0 - lbl Main
  await e;
  // TRAIL 0 - lbl Awake_e
  // TRAIL 0 - lbl And_chk
with
  // TRAIL 1 - lbl And_sub_2
  await A;
  // TRAIL 1 - lbl Awake_A_1
  emit e;
  // TRAIL 1 - lbl Emit_cont
  // TRAIL 1 - lbl And_chk
end
// TRAIL 0 - lbl And_out
await A;
// TRAIL 0 - lbl Awake_A_2

@@
